---
title: "Data Visualization"
author: "Heidi Golden"
date: "2023-03-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
See the BOOK: https://learning.nceas.ucsb.edu/2023-02-arctic/index.html

## Load Libraries
```{r Libraries}
  library(leaflet)# mapping
  library(dplyr)#
  library(tidyr)#
  library(ggplot2)#
  library(DT)# interactive tables
  library(scales) # install.packages("scales")#
```

## Read Data

```{r ReadData}
# Create a url object
data_url <- "https://knb.ecoinformatics.org/knb/d1/mn/v2/object/urn%3Auuid%3Af119a05b-bbe7-4aea-93c6-85434dcb1c5e"


# tryCatch function: tries to open escapement.csv, if error it will read from data_url
esc <- tryCatch(
    read.csv("data/escapement.csv"),
    error=function(cond) {
        message(paste("Escapement file does not seem to exist, so get it from the KNB."))
        esc <- read.csv(url(data_url, method = "libcurl"))
        return(esc)
    }
)

head(esc)
glimpse(esc)
```

## Challenge: Group by Year and calculate annual escapement
```{r ChallengeGroupCalc}

annual_esc <- esc %>% 
  separate(sampleDate, c("Year", "Month", "Day"), sep = "-") %>% # separates Date into columns
  mutate(Year = as.numeric(Year)) %>% # makes year numeric
  group_by(Species, SASAP.Region, Year) %>% # Groups into what you want to summarize by...
  summarize(escapement = sum(DailyCount)) %>% # Calculate the sum of yearly escapement
  filter(Species %in% c("Chinook", "Sockeye", "Chum", "Coho", "Pink")) # Choosing just salmonids

unique(annual_esc$Species)
```

## Plot static figures

Basic ggplot figure

```{r FirstPlot1}
ggplot(annual_esc, 
       aes(x = Species, y = escapement)) +
  geom_col()
```


```{r FirstPlot2}
ggplot(annual_esc, 
       aes(x = Species, y = escapement)) + #mapping aesthetics are variable that we are working with.
  geom_col(fill = "blue") #fill = blue goes with the geometry, so it goes here.
```

Plotting escapement for each species by region
```{r plottingByRegions3}
ggplot(annual_esc, 
       aes(x = Species, y = escapement, 
           fill = SASAP.Region)) + # Fill here works bcs its "by" data in the data frame.
  geom_col()
```

## 'ggplot' and the Pipe Operator

Plot Annual escapement from Kodiak by species:
You can pipe in data directly to ggplot and pipe adjustments, like filtering by Region...
```{r PipeOperator4}
annual_esc %>% 
  filter(SASAP.Region == "Kodiak") %>% # Can pipe in the data subset
  ggplot(aes(x = Year, y = escapement, color = Species)) +
  geom_line()+
  geom_point()
```

## Customize our plot using theme_

```{r CustomizingWithTheme5}
kodiak_esc <- annual_esc %>% 
  filter(SASAP.Region == "Kodiak") 

## == is the convention to look for a variable, == asks a question, is Region Kodiak?
## = is an assignment operator, it would assign the value Kodiak to the Region Column.

```

```{r NewPlot6}
ggplot(kodiak_esc,
       aes(x = Year,
           y = escapement,
           color = Species)) +
  geom_line()+
  geom_point()+
  ylab("Escapement")+
  ggtitle("Kodiak Salmon Escapement")+
  theme_bw() # Theme 

```


```{r NewPlotTheme7}
ggplot(kodiak_esc,
       aes(x = Year,
           y = escapement,
           color = Species)) +
  geom_line()+
  geom_point()+
  ylab("Escapement")+
  ggtitle("Kodiak Salmon Escapement")+
  theme_classic()

```

## Building on the code, Adding layers i.e. theme function

```{r NewPlotThemeFunction8}
ggplot(kodiak_esc,
       aes(x = Year,
           y = escapement,
           color = Species)) +
  geom_line()+
  geom_point()+
  ylab("Escapement")+
  ggtitle("Kodiak Salmon Escapement")+
  theme_bw()+
  theme(legend.position = "bottom",
        legend.title = element_blank()) # customizes components

## Theme function must come after the theme_...() choice.

## Nice!

```

## Saving a custom theme

```{r MyThemeCreate9}
my_theme <- theme_bw()+
  theme(legend.position = "bottom",
        legend.title = element_blank())
```

```{r MyThemeUse10}
ggplot(kodiak_esc,
       aes(x = Year,
           y = escapement,
           color = Species)) +
  geom_line()+
  geom_point()+
  scale_y_continuous(labels = comma)+
  ylab("Escapement")+
  ggtitle("Kodiak Salmon Escapement")+
  my_theme

# Adding scale_y_continuous changes the y axis
```

## Saving a plot

```{r SavingPlot11}
ggsave("kodiak_esc.png", width = 3, height = 3, units = "in")

```

## Creating Multiple Plots
Plot escapement throughout years for all of the Regions
```{r CreateMultiplePlots12}
ggplot(annual_esc, 
       aes(x = Year, 
           y = escapement, 
           color = Species)) + 
    geom_line() +
    geom_point() +
    scale_y_continuous(labels = comma) +
    facet_wrap(~SASAP.Region, scales = "free_y", ncol = 2) +
    ylab("Escapement") +
    my_theme

# facet_grid uses two variables
# To change the banner on each plot, try using different themes, i.e. theme_void or some other.

```