---
title: "Practice Data Cleaning and Manipulation"
author: "Heidi Golden"
date: "2023-03-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Chapter 12 in the Book
https://learning.nceas.ucsb.edu/2023-02-arctic/exercise-cleaning-and-manipulating-data.html

## Load Libraries
```{r LoadLivrary}
library(rvest)
library(dplyr)
library(readr)
library(janitor)

```


## Scraping bird code species name
```{r ScrapingBirdCode}
webpage <- read_html("https://www.pwrc.usgs.gov/bbl/manual/speclist.cfm")

tbls <- html_nodes(webpage, "table") %>% 
    html_table(fill = TRUE)

species <- tbls[[1]] %>% 
    clean_names() %>% 
    select(alpha_code, common_name) %>% 
    mutate(alpha_code = tolower(alpha_code))

```

## Load the data from ADC by copy url
```{r LoadFromURL}

pred <- read_csv("https://arcticdata.io/metacat/d1/mn/v2/object/urn%3Auuid%3A9ffec04c-7e2d-41dd-9e88-b6c2e8c4375e")

nest <- read_csv("https://arcticdata.io/metacat/d1/mn/v2/object/urn%3Auuid%3A982bd2fc-4edf-4da7-96ef-0d11b853102d")


```
## 3. Write function to translate spp codes into common names


```{r WriteFunction}
# checking what's in the nest spp names
unique(nest$species) # nest$species = species$alpha_code

code_names <- function(spp_df, code_df) {
  join_names <- left_join(spp_df, code_df, by = c("species" = "alpha_code"))
  return(join_names)
}

pred_names <- code_names(pred, species)
nest_names <- code_names(nest, species)

# join_names <- left_join(pred, species, by = join_by(species == alpha_code))
# OR left_join(pred, species, by = c("species" = "alpha_code"))


```

## 4. Calculate total number of pred. by year and spp

```{r SummariseNumPredByYr}
pred_sum <- pred_names %>% 
  group_by(year, common_name) %>% 
  summarise(total = sum(count, na.rm = TRUE), .groups = "drop") %>% 
  filter(!is.na(common_name)) %>% 
  arrange(desc(total))

```

## 4.5. Plot results from above

```{r PlotTotalPred}
# Can get the code from the BOOK



```

## 5. Calculate total number of egg predated by year and spp
## 6. Calculate total number of predators by year, join to summarize egg predation tabel and plot the results
