---
title: "Data Cleaning and Manipulation"
author: "Heidi Golden"
date: "2023-03-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Libraries

```{r}
library(dplyr)
library(tidyr)
library(readr)
```
Warnings: Masked functions - These base r functions are overwitten by dplyr.
#######################################
Attaching package: ‘dplyr’

The following objects are masked from ‘package:stats’:

    filter, lag

The following objects are masked from ‘package:base’:

    intersect, setdiff, setequal, union
##########################################
To unmask, dplry::filter(union) This 

## Load data
```{r}
catch_original <- read.csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1")
str(catch_original)
```
The select() command from dplyr package selects **columns**
the filter() command  ... selects **row**
```{r}
# Remove the All and notesRegCode columns from the catch_original data frame
catch_data <- catch_original %>% 
  select(-All, -notesRegCode) 
  #select(Region, Year, Chinook, Sockeye, Coho, Pink, Chum) # can reorder columns this way
head(catch_data)
glimpse(catch_data)

```
```{r}
# Mutate chinook class from character to numeric
catch_clean <- catch_data %>% 
  mutate(Chinook = as.numeric(Chinook))
 
# NAs introducted, Search for the NA value

i <- which(is.na(catch_clean$Chinook))
i # row 401 has NA

catch_data[i, ]
# located the odd value. It was a uppercase I
# Assuming the I should be a 1. But should go to original data to make sure.

catch_clean <- catch_data %>%
  mutate(Chinook = if_else(Chinook == "I", "1", Chinook)) %>% 
  mutate(Chinook = as.integer(Chinook))

glimpse(catch_clean)

```
## Change shape of the data using pivot longer and pivot wider
```{r}
#pivot longer practice
catch_long <- catch_clean %>% 
  pivot_longer(cols = -c(Region, Year), names_to = "species", values_to = "catch")

# Makes Species "long"
# cols = .. These are the columns that are retained as is.
# if year wasn't included in the cols = ... then the work year would appear in the species column with the year value.

# names_to is what you call the column of column names.
# values-to contains the values that were in the multiple species columns


head(catch_long)
```

```{r pivot_wide}
catch_wide <- catch_long %>% 
  pivot_wider(names_from = species, values_from = catch)

head(catch_wide)

```

## Practice Renaming
```{r Rename}
catch_long <- catch_long %>%
  rename(catch_thousands = catch)

head(catch_long)

# Best to rename using the header names, not the column position (i.e. "Year" vs [ ,2])
```
## Adding columns with mutate function
```{r}
catch_long <- catch_long %>%
  mutate(catch = catch_thousands * 1000)

head(catch_long)

# remove catch_thousands
catch_long <- catch_long %>%
  mutate(catch = catch_thousands * 1000) %>%
  select(-catch_thousands)

head(catch_long)
```

## Practice using group-by and summarise (or summarize) functions

```{r group-by-summarize}
mean_reagion <- catch_long %>%
  group_by(Region) %>%
  summarise(catch_mean = mean(catch))

head(mean_reagion)

```

```{r summariseCountRegion}
# summarize the number of observations (n) by region
# why two pipe opperators? Bcs we are doing two processes.

n_region <- catch_long %>% 
  group_by(Region) %>%
  summarize(n = n(),
            total_catch = sum(catch))

head(n_region)

```

```{r filter}
ALU_catch <- catch_long %>%
  filter(Region == c("ALU", "SSE") & species == c("Sockeye", "Pink"))

# == is same as %in%

head(ALU_catch)
```

## Arrange some Rows.
```{r Arrange}
mean_region_arrange <- catch_long %>% 
  group_by(Region) %>% 
  summarise(mean_catch = mean(catch)) %>% 
  arrange(-mean_catch) # can also use arrange(desc(mean_catch)), instead of negative sign

# arranges from smallest to largest
# NEGATIVE sign give largest to smallerst

head(mean_region_arrange)

# see entire data set...
mean_region_arrange
```

## Practice Joins
```{r loadDefs}
# loading in the region_defs .csv
region_defs <- read.csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.303.1") %>% 
    select(code, mgmtArea)
# just want to keep code and management area. No other columns.
# SWEET!

head(region_defs)
```

```{r Join}
catch_joined <- left_join(catch_long, region_defs, by = c("Region" = "code"))

head(catch_joined)
```

```{r Renaming}
region_defs <- region_defs %>% 
  rename(Region = code, Region_Name = mgmtArea)
# Rename columns we want to join by so they are the same.
# Rename switches after = to what's before =

catch_joined <- left_join(catch_long, region_defs, by = c("Region"))

head(catch_joined)

```

## Separate practice
```{r Separate}
sites_df <- data.frame(site = c("HAW-101",
                                "HAW-103",
                                "OAH-320",
                                "OAH-219",
                                "MAI-039"))


# Separates values in a column based on a character removes the hyphens
# Similar to text to columns in Excel !!! Yeah !!!
sites_df %>% 
  separate(site, c("island", "site_number"), "-") 

```

## Practice Uniting Columns
```{r Unite}
dates_df <- data.frame(year = c("1930",
                                "1930",
                                "1930"),
                       month = c("12",
                                "12",
                                "12"),
                       day = c("14",
                               "15",
                               "16"))

dates_df %>% 
  unite(date, year, month, day, sep = "-") %>%
  mutate(date = as.Date(date))

## Oh so TIDY !!
## can also use lubridate package posi

```

