---
title: "Geospatial Analysis with SF"
author: "Heidi Golden"
date: "2023-03-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
See the BOOK: https://learning.nceas.ucsb.edu/2023-02-arctic/geospatial-vector-analysis.html#visualize-sf-objects-with-leaflet

## Setup for geospatial analysis
```{r LoadLibraries}
library(sf) # Simple features - geospatial
library(ggplot2)
library(leaflet) # interactive maps
library(scales)
library(ggmap)
library(dplyr)
```
Plot 
```{r ReadingData}
## Read in shapefile using sf
ak_regions <- read_sf("data/ak_regions_simp.shp")

plot(ak_regions)
# these plots are need some work
# distorted

```
```{r CheckClass_ak_regions}
class(ak_regions)
# it's an sf object 
# but also has class tbl_df table and data frame.
# can use data frame functions

head(ak_regions)
# multipolygon with numeric mgmt areas, etc.
```
Coordinate Systems:
Unprojected = Geographic - 3D or globe (in angular units)
Projected defines where the data is on the earth (in linear units, usually meters)

 - Projection has lat and long represented as 2D on the earth
 - Shows how the data is drawn on a flat surface
 - For example how a peeled orange would look laid flat.


Halina's Notes:
CRS can be both unprojected and projected.
Unprojected CRS aka a geographic coordinate system (GCS)
- latitude and longitude references the earth as a 3D object
- tells us *where* something is located on the globe
- measured in angular units, i.e. degrees
Projected CRS or projected coordinated system (PCS) or projection (for short)
- latitude and longitude references the earth as a 2D object
- tells us *how* to draw or create the representation on a flat surface
- measured in linear units, i.e. meters

How to pick a geo ref system??
EPSG code = Coordinate Reference System (link in the BOOK) - provides a link to many, many codes
can search a region and finde all the codes for that region.
```{r CheckCoordSys}

st_crs(ak_regions) # Checks coord sys

# type of coord ref sys = GEOGCRS["WGS 84", which is an unprojected coordinate system.
# Need to assign the appropriate coordinate system
```

```{r trasformCRS}
# Pipe our data to the coordinate system
ak_regions_3338 <- ak_regions %>%
  st_transform(crs = 3338)
## 3338 is the EPSG code NAD83/Alaska Albers, which is a good one for AK

# Check if the transform worked
st_crs(ak_regions_3338)
# PROJCRS["NAD83 / Alaska Albers" - changed from GEO- to PROJ- CRS
#   BASEGEOGCRS["NAD83"
```

```{r ReplotProjected}
plot(ak_regions_3338)

# Much better now.
```

## st and the tidyverse

```{r sf_tidyverse}
ak_regions_3338 %>%
  select(region)

# geometry object will "stick" with the data - signals we are working w spatial data

```
```{r Filtering}
ak_regions_3338 %>%
  filter(region == "Southeast")


```

## Spatial Joins

```{r spatial_joins}
# read in population data
pop <- read.csv("data/alaska_population.csv")

# what to determine # of people per region
head(pop)
class(pop)
# need to turn pop into an sf object
```

### make 'pop' an sf object using 'st_as_sf()'

```{r MakePopsf}
# 4326 code is a default system
# Assume 4326 when we do not know what the coord system is.
# Very common is WGS 84

pop_4326 <- st_as_sf(pop, 
                  coords = c('lng', 'lat'),
                  crs = 4326,
                  remove = FALSE) # retains the original lat and lng columns !!!!!!!!

head(pop_4326)

# Added a new geometry column in addition to our lat, lng columns
# Now we can join
```

## assign each city to a region using 'st_join'

```{r WRongCOODs_NotMatchingFAILS}
pop_joined <- st_join(pop_4326, ak_regions_3338, join = st_within)

# ?st_join to see all different values for spatial joins that we can do.
# we want st_within


## Our EPSG codes do NOT MATCH. They must have the SAME Projection!
```

```{r CorrectingCOORDs}
pop_3338 <- st_transform(pop_4326, crs = 3338) # changing the coord system

st_crs(pop_3338)
```


```{r RedoJoin}
pop_joined <- st_join(pop_3338, ak_regions_3338, join = st_within)

head(pop_joined)
```
We assumed the populations were in 4326, and assumed they were in WGS, most GPS units use WGS83, but NAD83 is also used.

CHECK the metadata to see what the GPS units were taken in.
There are some packages that can check the coordinate system for you.

If you don't know the spatial metadata
see .prj file for projection.


## Calculate the Total Population per Region Using 'group-by' and 'summarize()'
```{r ChangeBackToDataframe}

pop_region <- pop_joined %>% 
  as.data.frame() %>% 
  group_by(region) %>% 
  summarise(total_pop = sum(population))

head(pop_region)

```


```{r}
# Left join the total population info with ...

pop_region_3338 <- left_join(ak_regions_3338, pop_region, by = "region")

## Joining with `by = join_by(region)`
```



```{r plotTotalPop_byregion}
plot(pop_region_3338["total_pop"])
```


```{r PlotAgain}
# using tidyverse on sf objects

pop_mgmt_3338 <- pop_region_3338 %>% 
  group_by(mgmt_area) %>% 
  summarize(total_pop = sum(total_pop), do_union = FALSE)

# 1. groups the population data by management area instead of region.
# 2. sums the populations by those management areas
# 3. do_union false retains the regional boundaries, so no merging of regional polygons.


# shows that we can use group_by and summarize over geometries.
# do_union if TRUE merges it merges the polygons (aka the geometries) 
# do_union FALSE keeps the polygons separate

plot(pop_mgmt_3338["total_pop"])

```


## Write spatial data as a file using 'write_sf()'

```{r WritingToFile}
# Creates a new .shp file
write_sf(pop_region_3338, "data/ak_regions_population.shp", delete_layer = TRUE) 

# delete_layer argument? If false, it will not let you overwrite the file.
# when true it allows you to overwrite the same layer.

################################################
# The Manual for using tidyverse with sf objects
#?st::tidyverse
```


## Visulaixe using ggplot

```{r VisWithGGPLOY}
ggplot(pop_region_3338) +
  geom_sf(aes(fill = total_pop)) +
  theme_bw() +
  labs(fill = "Total Population") +
  scale_fill_continuous(low = "khaki", high =  "firebrick", labels = comma) #scales package for colors

# comma is a function from scales
# scales is a companion to ggplot
# labels = percent
# labels = comma
# Also money and other things for the labels/legend
```

## Adding other shape files, i.e., river boundaries


```{r LoadRiversShapeFile}
rivers_3338 <- read_sf("data/ak_rivers_simp.shp")
st_crs(rivers_3338) # projection is in Albers


#################################
# There is a website where you can paste in the PROJCRS to find the correct projection
# crs suggest R package to confirm the projection
```

```{r PlotANotherPlot}
# Adding Rivers to the plot
# using three different spatial data sets

ggplot() +
  geom_sf(data = pop_region_3338, aes(fill = total_pop)) +
  geom_sf(data = rivers_3338, aes(size = StrOrder), color = "black") +
  geom_sf(data = pop_3338, aes(), size = .5) +
  scale_size(range = c(0.01, 0.2), guide = F) +
  theme_bw() +
  labs(fill = "Total Population") +
  scale_fill_continuous(low = "khaki", high =  "firebrick", labels = comma)


```

## Incorporate base maps into static maps using ggmap
```{r GGMAP}
pop_3857 <- pop_3338 %>%
  st_transform(crs = 3857) 

# can render as raster
# using stamen map time with CRS of 3857
# transform bounding box so it can be in 3857
```

Define bounding box
COOL
```{r FunctionToDefineBoundingBox}
# Define a function to fix the bbox to be in EPSG:3857
# See https://github.com/dkahle/ggmap/issues/160#issuecomment-397055208
ggmap_bbox_to_3857 <- function(map) {
  if (!inherits(map, "ggmap")) stop("map must be a ggmap object")
  # Extract the bounding box (in lat/lon) from the ggmap to a numeric vector, 
  # and set the names to what sf::st_bbox expects:
  map_bbox <- setNames(unlist(attr(map, "bb")), 
                       c("ymin", "xmin", "ymax", "xmax"))
  
  # Coonvert the bbox to an sf polygon, transform it to 3857, 
  # and convert back to a bbox (convoluted, but it works)
  bbox_3857 <- st_bbox(st_transform(st_as_sfc(st_bbox(map_bbox, crs = 4326)), 3857))
  
  # Overwrite the bbox of the ggmap object with the transformed coordinates 
  attr(map, "bb")$ll.lat <- bbox_3857["ymin"]
  attr(map, "bb")$ll.lon <- bbox_3857["xmin"]
  attr(map, "bb")$ur.lat <- bbox_3857["ymax"]
  attr(map, "bb")$ur.lon <- bbox_3857["xmax"]
  map
}
```

Define the bounding box for S AK:
```{r}
bbox <- c(-170, 52, -130, 64)   # This is roughly southern Alaska

# Get the basemap that we want using the bounding box defined above
ak_map <- get_stamenmap(bbox, zoom = 4)

# Use the function defined above o plot the map
ak_map_3857 <- ggmap_bbox_to_3857(ak_map)

```


```{r PlotTheMap}
# What is this.  inherit.aes FALSE??
ggmap(ak_map_3857) + 
  geom_sf(data = pop_3857, aes(color = population), inherit.aes = F) +
  scale_color_continuous(low = "khaki", high =  "firebrick", labels = comma)

#scale fill continuous ??


```
## Interactive Maps with leaflet
leaflet projects the data for us, but need to tell it the projection to use.
Shapefile must have 4326 code - WGS84
3338 code - AK Albers
```{r}
epsg3338 <- leaflet::leafletCRS(
  crsClass = "L.Proj.CRS",
  code = "EPSG:3338",
  proj4def =  "+proj=aea +lat_1=55 +lat_2=65 +lat_0=50 +lon_0=-154 +x_0=0 +y_0=0 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs",
  resolutions = 2^(16:7))
```


```{r TransformToUnprojected}
## Why does it need to be unprojected?

pop_region_4326 <- pop_region_3338 %>% 
  st_transform(crs = 4326)

#. EPSG:4326 - World Geodetic System 1984, used in GPS

```

Create simple interactive map:
```{r CreateInteractiveMap}
m <- leaflet(options = leafletOptions(crs = epsg3338)) %>%
        addPolygons(data = pop_region_4326, 
                    fillColor = "gray",
                    weight = 1)

m
```



```{r Tweak1}
pal <- colorNumeric(palette = "Reds", domain = pop_region_4326$total_pop)

m <- leaflet(options = leafletOptions(crs = epsg3338)) %>%
        addPolygons(data = pop_region_4326, 
                    fillColor = ~pal(total_pop),
                    weight = 1,
                    color = "black",
                    fillOpacity = 1,
                    label = ~region) %>% 
        addLegend(position = "bottomleft",
                  pal = pal,
                  values = range(pop_region_4326$total_pop),
                  title = "Total Population")

m
```


```{r Tweak2}
pal <- colorNumeric(palette = "Reds", domain = pop_region_4326$total_pop)

m <- leaflet(options = leafletOptions(crs = epsg3338)) %>%
        addPolygons(data = pop_region_4326, 
                    fillColor = ~pal(total_pop),
                    weight = 1,
                    color = "black",
                    fillOpacity = 1) %>% 
        addCircleMarkers(data = pop_4326,
                         lat = ~lat,
                         lng = ~lng,
                         radius = ~log(population/500), # arbitrary scaling
                         fillColor = "gray",
                         fillOpacity = 1,
                         weight = 0.25,
                         color = "black",
                         label = ~paste0(pop_4326$city, ", population ", comma(pop_4326$population))) %>%
        addLegend(position = "bottomleft",
                  pal = pal,
                  values = range(pop_region_4326$total_pop),
                  title = "Total Population")

m
```


## Links for More Info:
See the BOOK: https://learning.nceas.ucsb.edu/2023-02-arctic/geospatial-vector-analysis.html#visualize-sf-objects-with-leaflet
Chapter 13 for more info on spatial mapping

ALSO:
mapview package, too, gives you an instant interactive map to view.
Link: https://r-spatial.github.io/mapview/

This was in the Slack channel:
epsg3338 <- leaflet::leafletCRS(
  crsClass = "L.Proj.CRS",
  code = "EPSG:3338",
  proj4def =  "+proj=aea +lat_1=55 +lat_2=65 +lat_0=50 +lon_0=-154 +x_0=0 +y_0=0 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs",
  resolutions = 2^(16:7))
