---
title: "Practice Data"
author: "Heidi Golden"
date: "2023-02-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# PLACE Packages here!!

library(readr) #loads library to working space
```
See the BOOK: https://learning.nceas.ucsb.edu/2023-02-arctic/index.html


FIRST: Create an outline -- what we want to do with this analysis

## Introduction

This data set contains North Pole Environmental Observatory Bottle Chemisty. Here's a link to the data:
[LINK](https://arcticdata.io/catalog/view/doi:10.18739/A25T3FZ8X)

## Read in the data
```{r}
bg_chem <- read_csv("data/BGchem2008data.csv")
```

```{r dataexploration, echo=FALSE}
colnames(bg_chem)

str(bg_chem)

head(bg_chem)
```


## Analysis

### Calculate summary statistics
```{r stats}

nitrate <- mean(bg_chem$NO3)
nitrite <- mean(bg_chem$NO2)
amm <- mean(bg_chem$NH4)
phos <- mean(bg_chem$P)

```


### Calculate mean Redfield ratio

```{r RedfieldRatio}
ratio <- (nitrate + nitrite + amm)/phos
# Now we can write text and call these numbers within the text..
```

The Redfield ratio in this dataset is approximately `r round(ratio)`


### Plot the Redfiled ratios

```{r PlotTheData, echo=FALSE}
plot(bg_chem$P, bg_chem$NO2 + bg_chem$NO3 + bg_chem$NH4)
```

